---
title: "Ovarian Cancer proteomics - prediction¨"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


## Libraries 
```{r}
#renv::install(c("hadley/emo", "heike/extracat"))
pkgs <- c(
  "AmesHousing",
  "AppliedPredictiveModeling",
  "bookdown",
  "broom",
  "caret",
  "caretEnsemble",
  "cluster",
  "cowplot",
  "DALEX",
  "data.table",
  "doParallel",
  "dplyr",
  "janitor",
  "dslabs",
  "e1071",
  "earth",
 # "emo",
 # "extracat",
  "factoextra",
  "foreach",
  "forecast",
  "ggbeeswarm",
  "ggmap",
  "ggplot2",
  "ggplotify",
  "gbm",
  "glmnet",
  "gridExtra",
  "h2o",
  "HDclassif",
  "iml",
  "ipred",
  "kableExtra",
  "keras",
  "kernlab",
  "knitr",
  "lime",
  "markdown",
  "MASS",
  "Matrix",
  "mclust",
  "mlbench",
  "NbClust",
  "pBrackets",
  "pcadapt",
  "pdp",
  "plotROC",
  "pls",
  "pROC",
  "purrr",
  "ranger",
  "readr",
  "recipes",
  "reshape2",
  "ROCR",
  "rpart",
  "rpart.plot",
  "rsample",
  "scales",
  "sparsepca",
  "stringr",
  "subsemble",
  "SuperLearner",
  "tfruns",
  "tfestimators",
  "tidyr",
  "vip",
  "visdat",
  "xgboost",
  "yardstick"
)

renv::install(pkgs)
```

```{r}
library(tidyverse)
library(h2o)
```



```{r}
sumExpObj_overall <- readRDS( "../data/rawdata/data_sumExpfil_wntr_new_vars.rds")
```

 - Data prparation: pre-processing the feature and target variables, 
 - minimizing data leakage (Section 3.8.2), ??
 - Tuning hyperparameters
 - Assessing model performance. 

## Data preparation

The normalized data is used for machine learning approach. Initially, 1)
the Malignant vs. benign from the "type" variable and 2) categories 2 vs
zero from the "type_1\_2" variable are the focus. 3) Re-categorise BL
samples to B (Benign category)

This is due to sample sizes in these subgroups.

## Remove correlated fatures

```{r}
#sumExpObj_overall <- sumExpObj_overall[, sumExpObj_overall$tumor_type %nin% c("Other", "BL")]
```

## Filter unwanted samples from the data and fix lables

```{r}

library(Hmisc)
library(ggplot2)
sumExpObj_overall_ml_type <- sumExpObj_overall[, sumExpObj_overall$tumor_type %nin% c( "Other" , "H")]
sumExpObj_overall_ml_type$tumor_type <- droplevels(sumExpObj_overall_ml_type$tumor_type)
sumExpObj_overall_ml_type$tumor_type_2 <- ifelse(sumExpObj_overall_ml_type$tumor_type == "B" | sumExpObj_overall_ml_type$tumor_type == "BL",  "benign","malignant" )

sumExpObj_overall_ml_type12 <- sumExpObj_overall[, sumExpObj_overall$Type_1_2 != c("1") ]
sumExpObj_overall_ml_type12$Type_1_2  <- sumExpObj_overall_ml_type12$Type_1_2 |> droplevels() |> as.numeric()
sumExpObj_overall_ml_type12$Type_1_2  <- (sumExpObj_overall_ml_type12$Type_1_2  - 1) 

rowData(sumExpObj_overall_ml_type12)$protein <- str_split(rownames(sumExpObj_overall_ml_type12), pattern = ";", simplify = TRUE)[,1] |> make.unique()
```



## Then we remove the highly correlated features from the data

```{r}
library(caret)
nzv <- nearZeroVar(t(assays(sumExpObj_overall)$norm_imputed), saveMetrics= TRUE)
descrCor <-  cor(t(assays(sumExpObj_overall)$norm_imputed), use = "na.or.complete" )
highCorr <- sum(abs(descrCor[upper.tri(descrCor)]) > .95)
highlyCorDescr <- findCorrelation(descrCor, cutoff = .90)
low_var <- which(rowData(sumExpObj_overall)$rowVars <)
#sumExpObj_overall_ml_type <- sumExpObj_overall_ml_type[-highlyCorDescr,]
sumExpObj_overall_ml_type12 <- sumExpObj_overall[-highlyCorDescr,]
sumExpObj_overall_ml_type12 <- sumExpObj_overall[-highlyCorDescr,]
data_type_12 <- data.frame(sumExpObj_overall_ml_type12$tumor_type_new,  t(assays(sumExpObj_overall_ml_type12)$norm_imputed) )
colnames(data_type_12)[1] <- "label"
data_type_12$label <- as.factor(data_type_12$label)
```

## Train test split and scaling of data

```{r}
set.seed(123)


trainIndex <- createDataPartition(data_type_12$label, p = .9, 
                                  list = FALSE, 
                                  times = 1)
test <- preProcess(data_type_12[-1])

train_prot <- data_type_12[trainIndex,]
train_prot_scaled <- predict(test, train_prot[, -1])
#train_prot_scaled <- data.frame(train_prot$label)
train_prot_all <- cbind(label=train_prot$label, train_prot_scaled) |> as.data.frame()
test_prot <- data_type_12[-trainIndex,]
test_prot_scaled <- predict(test, test_prot[,-1])
test_prot_all <- cbind(label = test_prot$label, test_prot_scaled) |> as.data.frame()
```


## Build models

### With h2O and DALEX

```{r}
#h2o.removeAll() # Clean up. Just in case H2O was already running
h2o.init(nthreads = -1, max_mem_size="10G")  # Start an H2O cluster with all threads available
# convert to h2o object
df.h2o <- as.h2o(data_type_12)

# create train, validation, and test splits
#set.seed(123)
splits <- h2o.splitFrame(df.h2o, ratios = 0.8, destination_frames = c("train","test"), seed = 1234)
names(splits) <- c("train","test")

# variable names for resonse & features
response <- "label"
features <- setdiff(names(data_type_12), response) 
```


```{r}
n_features <- length(features)
hyper_grid_1 <- list(ntrees = seq(50, 200, by = 20),
                     mtries = floor(n_features * c(0.025, 0.040 , .05,0.075, 0.1, 0.15, 0.2)),
                    min_rows = c(1, 3, 5, 10),
                  max_depth = c(1, 5,10, 20, 25,30, 35),
                  sample_rate = seq(0.7, 0.85, by = 0.05),
                  nbins = seq(20, 80, 10)
                  
)

 search_criteria = list(strategy = "RandomDiscrete", 
                        max_models = 100, 
                        max_runtime_secs = 1000 * 60
                        )
 
grid_crtsn <- h2o.grid( algorithm = "randomForest",
                        x = features,
                        y= response,
                        grid_id = "rf_grid2",
                        nfolds= 3,
                        stopping_metric = "AUC", 
                        training_frame = splits$train,
                        stopping_rounds = 5,         
                        stopping_tolerance = 0.005,
                        seed = 1234,
                        hyper_params = hyper_grid_1,
                       search_criteria = search_criteria
    
)
```


```{r}
best_model <-  h2o.get_best_model()
```


```{r}
grid_perf <- h2o.getGrid(grid_id = "rf_grid2",
                         sort_by = "AUC",
                         decreasing = T)
grid_perf@summary_table
best_model <-  h2o.get_best_model(grid_perf)
h2o.varimp(best_model)
pred <- h2o.predict(object = best_model, newdata = splits$test)
```


```{r}
# random forest model
rf <- h2o.randomForest(y="label",
  training_frame = splits$train,
  validation_frame = splits$valid,
  ntrees = 1000,
  stopping_metric = "AUC",    
  stopping_rounds = 10,         
  stopping_tolerance = 0.005,
  seed = 123
  )

```

## Logistic regression

## Random forrest 
1. The number of trees in the forest
2. The number of features to consider at any given split:  
3. The complexity of each tree
4. The sampling scheme
5. The splitting rule to use during tree construction

and (2) typically have the largest impact on predictive accuracy and should always be tuned. (3) and (4) tend to have marginal impact on predictive accuracy but are still worth exploring. They also have the ability to influence computational efficiency. (5) tends to have the smallest impact on predictive accuracy and is used primarily to increase computational efficiency.

```{r}
# train a default random forest model
prot_rf1 <- ranger(
  label ~ ., 
  data = train_prot_all,
  mtry = floor(n_features / 3),
  respect.unordered.factors = "ignore",
  seed = 123
)

# get OOB RMSE
(default_rmse <- sqrt(prot_rf1$prediction.error))
```

```{r}
library(ranger)
n_features <- length(setdiff(names(train_prot_all), "label"))
# create hyperparameter grid
hyper_grid <- expand.grid(
  mtry = floor(n_features * c(0.25, .05,0.075, 0.1, .15, 0.2 ,.25, .333, .4)),
  min.node.size = c(1,2, 3, 4, 5, 10), 
  replace = FALSE,                               
  sample.fraction = c(.5, .63,  0.7, .8),                       
  rmse = NA 
)

sapply(hyper_grid, length ) %>% prod()
# execute full cartesian grid search
for(i in seq_len(nrow(hyper_grid))) {
  # fit model for ith hyperparameter combination
  fit <- ranger(
    formula         = label ~ ., 
    data            = train_prot_all, 
    num.trees       = n_features * 10,
    mtry            = hyper_grid$mtry[i],
    min.node.size   = hyper_grid$min.node.size[i],
    replace         = hyper_grid$replace[i],
    sample.fraction = hyper_grid$sample.fraction[i],
    verbose         = FALSE,
    seed            = 123,
    respect.unordered.factors = 'ignore',
  )
  # export OOB error 
  hyper_grid$rmse[i] <- sqrt(fit$prediction.error)
}

# assess top 10 models
hyper_grid %>%
  arrange(rmse) %>%
  mutate(perc_gain = (default_rmse - rmse) / default_rmse * 100) %>%
  head(20)
```





```{r}
library(h2o)
h2o.no_progress()
h2o.init(max_mem_size = "8g")
```

```{r}
# convert training data to h2o object
train_h2o <- as.h2o(train_prot_all)

# set the response column to label
response <- "label"

# set the predictor names
predictors <- setdiff(colnames(train_prot_all), response)
```


```{r}
h2o_rf1 <- h2o.randomForest(
    x = predictors, 
    y = response,
    training_frame = train_h2o, 
    ntrees = n_features * 10,
    seed = 123
)

h2o_rf1

```

```{r}
# hyperparameter grid
hyper_grid <- list(
  mtries = floor(n_features * c(0.25, .05,0.075, 0.1, .15, 0.2 ,.25, .333, .4)),
  min_rows = c(1, 3, 5, 10),
  max_depth = c(1, 5,10, 20, 30),
  sample_rate = c( .632, .70, .80)
)

# random grid search strategy
search_criteria <- list(
  strategy = "RandomDiscrete",
  stopping_metric = "auc",
  stopping_tolerance = 0.001,   # stop if improvement is < 0.1%
  stopping_rounds = 10,         # over the last 10 models
  max_runtime_secs = 60*5      # or stop search after 5 min.
)
```


```{r}
# perform grid search 
random_grid <- h2o.grid(
  algorithm = "randomForest",
  grid_id = "hyper_grid",
  x = predictors, 
  y = response, 
  training_frame = train_h2o,
  hyper_params = hyper_grid,
  ntrees = n_features * 10,
  seed = 123,
  stopping_metric = "auc",   
  stopping_rounds = 10,           # stop if last 10 trees added 
  stopping_tolerance = 0.005,     # don't improve RMSE by 0.5%
  search_criteria = search_criteria
)
```


```{r}
# collect the results and sort by our model performance metric 
# of choice
random_grid_perf <- h2o.getGrid(
  grid_id = "hyper_grid", 
  sort_by = "auc", 
  decreasing = TRUE
)
random_grid_perf 
```


```{r}
library(ranger)
# re-run model with impurity-based variable importance
rf_impurity <- ranger(
  formula = label ~ ., 
  data = train_prot_all, 
  num.trees = 40,
  mtry = 17,
  min.node.size = 1,
  sample.fraction = .7,
  replace = FALSE,
  importance = "impurity",
  respect.unordered.factors = "ignore",
  verbose = FALSE,
  seed  = 123
)

# re-run model with permutation-based variable importance
rf_permutation <- ranger(
  formula = label ~ ., 
  data = train_prot_all, 
  num.trees = 40,
  mtry = 17,
  min.node.size = 1,
  sample.fraction = 0.7,
  replace = FALSE,
  importance = "permutation",
  respect.unordered.factors = "ignore",
  verbose = FALSE,
  seed  = 123
)
```


```{r}
rf_impurity
```


```{r}
p1 <- vip::vip(rf_impurity, num_features = 25, bar = FALSE)
p2 <- vip::vip(rf_permutation, num_features = 25, bar = FALSE)

gridExtra::grid.arrange(p1, p2, nrow = 1)
```


```{r}
explainer_ranger <- explain(rf_impurity, data = train_prot_all[, -1], (as.numeric(train_prot_all$label)) - 1)
```

```{r}
# Compute predictions
pred1 <- predict(rf_impurity, test_prot_all[, -1]) %>% as.vector()

pred2 <- predict(rf_permutation, test_prot_all[, -1]) %>% as.vector()
```



```{r, message=FALSE, warning=FALSE}
qplot(pred1$predictions) + theme_minimal()
```

```{r}
# Review AUC - Area Under Curve
# 
# prediction3 <- as.numeric(pred1$predictions > 0.5)
# prediction4 <- as.numeric(pred2$predictions > 0.5)
confusionMatrix(pred1$predictions,test_prot_all$label)
confusionMatrix(pred2$predictions, test_prot_all$label)

```


```{r}
plot(pROC::roc(response = test_prot_all$label,
               predictor = as.numeric(pred2$predictions),
               levels=c(0, 1)),  lwd=2) 
```



## Feature selection
```{r}
library(SuperLearner)
library("arm")
set.seed(123)
# Fit the SuperLearner.
# We need to use list() instead of c().
SL.library <- c("SL.ranger", "SL.xgboost", "SL.svm" )
cv_sl = SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        #cvControl = list(V = 10),
                        #parallel = "multicore",
                        #method = "AUC",
                        #SL.library =  list("SL.randomForest",  c("SL.randomForest", "screen.corP"
                        SL.library =  SL.library)
```


```{r}
cv_sl

```

```{r}
preds <- predict.SuperLearner(cv_sl, test_prot_all[, -1])

prediction <- as.numeric(preds$pred > 0.5)
cm <- confusionMatrix(as.factor(prediction), as.factor(as.numeric(test_prot_all$label)-1))

cm
```
```{r}
set.seed(167)

# Fit the SuperLearner.
# We need to use list() instead of c().
 SL.library <- list("SL.randomForest",  c("SL.randomForest", "screen.corP"), "SL.ranger", c("SL.ranger", "screen.corP"),"SL.xgboost", c("SL.xgboost", "screen.corP"), "SL.svm" ,  c("SL.svm", "screen.corP")) 

cv_sl1 = CV.SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        cvControl = list(V = 10),
                        parallel = "multicore", SL.library =  SL.library
                        #list("SL.randomForest",  c("SL.randomForest", "screen.corP"
)

summary(cv_sl1)

cv_sl2 = CV.SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        cvControl = list(V = 10),
                        parallel = "multicore",
                        method = "method.AUC",
                        SL.library =   SL.library)

"method.NNloglik"

cv_sl3 = CV.SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        cvControl = list(V = 10),
                        parallel = "multicore",
                        method = "method.NNloglik",
                        SL.library =   SL.library)
```


```{r}
plot(cv_sl1)
```
```{r}
summary(cv_sl2)
```


```{r}
plot(cv_sl2)
```

```{r}
summary(cv_sl3)
```


It turns out that *rnadomForrest* with *corP* and  and *ranger* with *corP* parameters perform rather well as the confidence intervals around errors are smaller.
The *svm* with *corP* also performs well but with greater error. We will use ranger and svm for our analyses.

## Model tuning

To tune the model, we will use the create.learner function. 
Parameters to optimise for ranger are 
1) number of trees
2) mtry 



```{r}
learner_1 <- create.Learner("SL.ranger", params = list(num.trees = 2000))
SL.library <- list( c("SL.randomForest", "screen.corP"), "SL.ranger", c("SL.ranger", "screen.corP"), learner_1$names)
cv_sl4 = CV.SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        V = 10,
                        method = "method.AUC",
                        SL.library = SL.library)
summary(cv_sl4)
```

## Tune the *mtry* parameter.

```{r}
mtry_seq <-floor(sqrt(ncol(train_prot_all[ ,-1]))* c(0.15, 0.25, 0.5,0.7, 0.9, 1, 1.5, 2, 2.5, 3))
learners = create.Learner("SL.ranger", tune = list(mtry = mtry_seq),params = list(num.trees = 2000))

SL.library <- list(  learner_1$names)
cv_sl5 = CV.SuperLearner(Y = as.numeric(train_prot_all$label)-1, X = as.data.frame(train_prot_all[ ,-1]), family = binomial(),
                        V = 10,
                        method = "method.AUC",
                        SL.library = SL.library)
summary(cv_sl5)
```


```{r}
# Print the highest and lowest predicted sales price
paste("Observation", which.max(predictions), 
      "has a predicted sale price of", scales::dollar(max(predictions))) 
## [1] "Observation 1825 has a predicted sale price of $663,136"
paste("Observation", which.min(predictions), 
      "has a predicted sale price of", scales::dollar(min(predictions)))  
## [1] "Observation 139 has a predicted sale price of $47,245.45"

# Grab feature values for observations with min/max predicted sales price
high_ob <- as.data.frame(train_h2o)[which.max(predictions), ] %>% select(-Sale_Price)
low_ob  <- as.data.frame(train_h2o)[which.min(predictions), ] %>% select(-Sale_Price)
```


## Support vector machine


```{r}
# preProcValues <- preProcess(train_prot$data, method = c("center", "scale"))
# 
# trainTransformed <- predict(preProcValues, train_prot$data)
# preProcValues2 <- preProcess(test_prot$data, method = c("center", "scale"))
# 
# testTransformed <- predict(preProcValues2, test_prot$data)


xgb_train = xgb.DMatrix(data = train_prot$data, label = train_prot$label)
output_vector = train_prot$label == "1"
xgb_test = xgb.DMatrix(data = test_prot$data, label = test_prot$label)
```

```{r}
require(xgboost)
require(Matrix)
require(data.table)

watchlist = list(train=xgb_train, test=xgb_test)

#fit XGBoost model and display training and testing data at each round
#model = xgb.train(data = xgb_train, max.depth = 3, watchlist=watchlist, nrounds = 70)

final = xgboost(data = xgb_train, eta = 1, max.depth = 10, nrounds = 56, verbose = 0, objective = "binary:logistic", eval_metric = 'auc', prediction = T)
```

```{r}
importance <- xgb.importance(feature_names = final$feature_names, model = final)
head(importance)
```

```{r}
xgb.plot.importance(importance_matrix = importance)

```

```{r}
c2 <- cor(train_prot_all[,"Q13421"], output_vector, use = "complete.obs")
print(c2)
```

```{r}
pred <- predict(final, xgb_test)

# size of the prediction vector
print(length(pred))
```

```{r}
print(head(pred))
```

```{r}
prediction <- as.numeric(pred > 0.5)
print(head(prediction))
```

```{r}
library(pROC)
```

```{r}
plot(pROC::roc(response = test_prot$label,
               predictor = prediction,
               levels=c(0, 1)),
     lwd=1.5) 
```

```{r}
err <- mean(as.numeric(pred > 0.5) != test_prot$label)
print(paste("test-error=", err))
```

```{r}
confusionMatrix(as.factor(prediction), as.factor(test_prot$label))
```

```{r}
it = which.max(final$evaluation_log$test_auc_mean)
best.iter = final$evaluation_log$iter[it]
```

```{r}
#getwd()
Sys.setenv(RETICULATE_PYTHON = "penv/bin/python")
```

```{python}
import pandas as pd
from scipy import stats
from sklearn.preprocessing import StandardScaler
from sklearn.model_selection import RandomizedSearchCV, KFold
from sklearn.model_selection import train_test_split
from sklearn.metrics import classification_report
from xgboost import XGBClassifier
```

```{r}
features_all <- assays(sumExpObj_overall)$batch_rm_l_raw_loess |> as.data.frame() |> rownames_to_column(var = "rowname")
samples_data <- colData(sumExpObj_overall)  |> as.data.frame() |> rownames_to_column(var = "sampleID")

```

```{r}
nzv <- nearZeroVar(features_all, saveMetrics= TRUE)
descrCor <-  cor(t(assays(sumExpObj_overall)$batch_rm_l_raw_loess), use = "p")
highCorr <- sum(abs(descrCor[upper.tri(descrCor)]) > .95)
highlyCorDescr <- findCorrelation(descrCor, cutoff = .9)
highlyCorDescr <- findCorrelation(descrCor, cutoff = .75)
```

```{python}
features_all_py = r.features_all
samples_data_py = r.samples_data
```

```{r}
2^1.26
```


## References

https://bradleyboehmke.github.io/HOML/