---
title: | 
  | Section: 
  | Ovarian Proteomics: QC on the proteomics data
output:
  bookdown::html_document2:
    theme: journal
    highlight: kate
    df_print: paged
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
    code_folding: hide
    css: "style.css"
date: "`r format(Sys.time(), '%d %B, %Y')`"    
bibliography: references.bib 
editor_options: 
  chunk_output_type: console
---

```{r knitr, include=FALSE}
# knitr
knitr::opts_chunk$set(echo = TRUE, 
                      collapse = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      cache = FALSE, 
                      cache.path = "../cache/", 
                      fig.height = 5, 
                      fig.width = 6,
                      eval=TRUE)
```

# Purpose 
The main aim of the analyses is to QC the protein data and identify any outlier samples and batch effects


## Issues to discuss


 - Missing values
 - Normalization 
 - Variables to regress, time, death, ???

 

## QC of data

```{r setup1, echo=FALSE ,include=FALSE}
# import libraries and data
renv::init()
library(tidyverse)
library(reshape2)
library(magrittr)
library(readxl)
library(fuzzyjoin)
library(limma)
library(DBI)
library(VIM)
library(factoextra)
library(FactoMineR)
library(mice)
library(SummarizedExperiment)


library(RColorBrewer)
library(GGally)
library(ggiraphExtra)
library(knitr)
library(kableExtra)
library(plotly)
library(RankAggreg)
library(DT)
library(ggVennDiagram)
```


```{r setup1, echo=FALSE ,include=FALSE}
nn_sp_report <- read_tsv("../20210411_125835_Niklas N direct DIA_Report.xls", 
                       col_names = TRUE
                        )

sample_prep <- readxl::read_xlsx("../2021-03-05 Niklas Niehlsen with sample preparation dates.xlsx")
sample_clin <- readxl::read_xlsx("../Region.Skane.TTM1.Biobank_Chsoen_for_currentanalysis.xlsx") # The clinical data provided. The data contained wrong sample identifiers and one 0 was missing ans well as Malmo samples were coded as MAMO in the clinical data while the label was coded as MAL in the protein data.
# set seed for imputations
set.seed(20210527)

head(nn_sp_report)
```

```{r}
#nn_sp_report_unq_pep <- '.*?PG.Quantity'
nn_sp_report_un_pep <- nn_sp_report %>% distinct(PG.ProteinGroups,.keep_all = TRUE) %>% ungroup() # remove repeated values. The peptide level data is not provided, instead protein intensity was repeated for each protein.

nn_sp_report_un_pep <- nn_sp_report_un_pep %>% mutate(ID = str_split(PG.ProteinGroups, simplify = TRUE, pattern = ";")[,1], .before=1) # Add protein ID column (First in the protein list)

nn_sp_report_un_pept <- nn_sp_report_un_pep %>% select(contains(c("ID",'PG.Quantity'))) # Select the columns containing protein quantity

### Prepare the sample meta data by combining sample preparation sheet, annotations extracted from the protein column names and the clinical data provided. 
samples <- colnames(nn_sp_report_un_pept)[-1] # 
col_data <- data.frame(time = str_match(samples, "-[0-9]{2}-"),  sample =  str_split(str_split(samples, ".htrms.", simplify = TRUE)[,1], pattern = " ", simplify = TRUE)[,2] )
col_data <- col_data %>%  mutate(screening_number = str_sub(sample, start = 1, end = 9)  )

col_data <- left_join(col_data, sample_prep, by = c("sample"="Sample ID"))
rownames(col_data) <- col_data$sample
sample_clin <- as.data.frame(sample_clin[, -1])
col_data <- left_join(col_data, sample_clin, by= c("screening_number"="screening_number"), copy = TRUE)
rownames(col_data) <- col_data$sample

### Prepare the rowData from the row annotations
row_data <- nn_sp_report_un_pep[, 1:6]
rownames(row_data) <- nn_sp_report_un_pept$ID

raw_data <- as.matrix(nn_sp_report_un_pept[, samples])
rownames(raw_data) <- nn_sp_report_un_pept$ID

raw_data <- matrix(nn_sp_report_un_pept[, samples], dimnames = list(nn_sp_report_un_pept$ID, col_data$sample))
is.na(raw_data) %>% rowSums()
row_data <- row_data %>% mutate(ID = str_split(PG.ProteinGroups, simplify = TRUE, pattern = ";")[,1])
raw_data[is.nan(raw_data)] <- NA

sumExpObj <- SummarizedExperiment::SummarizedExperiment(assays = list(raw_data = raw_data),
    colData=col_data,
    rowData=row_data
)
saveRDS(sumExpObj, "../rawaDataSe.rds")
```


## Analyze missingness 

```{r}
sumExpObj <- readRDS("../rawaDataSe.rds")
#setwd("../missing/")
colData(sumExpObj)
colData(sumExpObj)$timepoint <- colData(sumExpObj)$time %>% as.factor() %>% as.integer() %>% as.character()
colnames(sumExpObj) <- paste0(colData(sumExpObj)$screening_number, "_", colData(sumExpObj)$timepoint )

## Analyze missingness and Impute for PCA etc
#Data_se <- readRDS("../Data/data_sumExp.rds")

missing <-mice::md.pattern(assays(sumExpObj)$raw_data)

png("../missing/missingProt.png", res = 300, width = 12000, height = 8000)
mice_prot <- aggr(t(assays(sumExpObj)$raw_data), col=c('navyblue','yellow'),
 numbers=TRUE, sortVars=TRUE,
 labels=sumExpObj$screening_number, cex.axis=1,
 gap=1, ylab=c("Missing data","Pattern"))
dev.off()

mice_prot$missings$Count %>% hist(breaks=200)


png("../missing/missingSamples.png", res = 300, width = 8000, height = 3000)
mice_sample <- aggr(assays(sumExpObj)$raw_data, col=c('navyblue','yellow'),
 numbers=TRUE, sortVars=TRUE,
 labels=sumExpObj$screening_number, cex.axis=.4,
 gap=2, ylab=c("Missing data","Pattern"))
dev.off()

sumExpObj_filt <- sumExpObj[mice_prot$missings[mice_prot$missing$Count <30,]$Variable,]

imp <- mice(assays(sumExpObj_filt)$raw_data, maxit = 0)

predM = imp$predictorMatrix
meth = imp$method
set.seed(123)
imp2 <- mice(assays(sumExpObj_filt)$raw_data, m=3, method = "cart")

imp3 <-  mice::complete(imp2, 1)

assays(sumExpObj_filt)$imputed_raw <- imp3

#saveRDS(sumExpObj_filt, "../data_sumExpfilImp.rds")
sumExpObj <- readRDS("../data_sumExpfilImp.rds")
```

```{r}
clin_vars <- c("Batch"  ,"Sample.preparation.date", "time","randomisation_code", "dead24",	"dead48",	"dead72","binarysecout180daycpcscore", "b_age",	"b_sex", "p_ca_to_rosc_min", "a_shock_on_admission", "shockable", "awakeday1",	"awakeday2",	"awakeday3")
inflam <- c("il6_1",	"il6_2",	"il6_3", "copeptin_1",	"copeptin_2",	"copeptin_3",	"crea_1",	"crea_2",	"crea_3",	"crp_1",	"crp_2",	"crp_3")
tech_var <- c("Batch"  ,"Sample.preparation.date" )
colData(sumExpObj)[, clin_vars]
```


## PCA

```{r}
library()
library("FactoMineR")

res.pca <- PCA(t(log2(assays(sumExpObj_filt)$imputed_raw)),  graph = TRUE)
```


```{r}
fviz_screeplot(res.pca, addlabels = TRUE, ylim = c(0, 50))
```

```{r}
rowData(sumExpObj)$na_per <- (rowSums(is.na(assay(sumExpObj)))/dim(sumExpObj)[2])*100
sumExpObj_filt <- sumExpObj[rowData(sumExpObj)$na_per <30 , ]
assay(sumExpObj_filt, "log2plus1") <- log2(assay(sumExpObj))+1

```

```{r}
limma::plotDensities(log2(assays(sumExpObj_filt)$imputed_raw), sumExpObj_filt$time)
```

```{r}

limma::plotDensities(log2(assays(sumExpObj_filt)$raw_data), sumExpObj_filt$time)

```

```{r}

assays(sumExpObj_filt)$vsn <- limma::normalizeVSN(assays(sumExpObj_filt)$raw_data)

limma::plotDensities(assays(sumExpObj_filt)$vsn, sumExpObj_filt$time)
```

```{r}
vsn::meanSdPlot(assays(sumExpObj_filt)$vsn)
```


```{r}

assays(sumExpObj_filt)$loess <- limma::normalizeCyclicLoess(log2(assays(sumExpObj_filt)$raw_data), iterations = 3) # Normalize using Loess on log2 trnasformed data
limma::plotDensities(assays(sumExpObj_filt)$loess, sumExpObj_filt$time)
```

```{r}


#assays(sumExpObj_filt)$batch_loess <- limma::removeBatchEffect(log2(assays(sumExpObj_filt)$loess), batch = sumExpObj_filt$Batch)
limma::plotDensities(assays(sumExpObj_filt)$loess, sumExpObj_filt$time)
saveRDS(sumExpObj_filt, "../data_sumExpfilImp.rds")
```

```{r}
res.pca_loess <- PCA(t(assays(sumExpObj_filt)$loess),  graph = TRUE)
```

```{r}
exp_var <- fviz_screeplot(res.pca_loess, addlabels = TRUE, ylim = c(0, 12))
ggsave(exp_var,
  file = paste("../",
    "explained_var", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)
```

```{r}
# scores plot (interactive)
ind <- get_pca_ind(res.pca_loess)
ind
fviz_pca_ind(res.pca, col.ind = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE # Avoid text overlapping (slow if many points)
             )
pca.frame <- data.frame(PC1 = round(res.pca$ind$coord[,1], 2), 
                        PC2 = round(res.pca$ind$coord[,2], 2), 
                        PC3 = round(res.pca$ind$coord[,3], 2), 
                        PC4 = round(res.pca$ind$coord[,4], 2),
                        PC5 = round(res.pca$ind$coord[,5], 2),
                        group = factor(data.naomit$Group, labels = c("Low", "High")),
                        SCANB = data.naomit$SCANB)

  
pca.var <- res.pca$eig[,3] %>%
  round(2)
  

```

### PCA analyses

Principle component analyses was used to perform exploratory analyses of the proteomics data. Explained variance by each principle component was calculated using the output from prcomp´ function in R. Correlation between principle components and phenotype (clinical) variables was calculated using ´lm´ function in R and "adj.r.squared" values were extracted from the summary statistics produced by the ´lm´ function.

```{r, eval = FALSE,include=FALSE,  echo = FALSE, fig.width=10, fig.height=18, warning = FALSE, message= FALSE}
is.na(assays(Data_se_comp)$imputed) %>% sum()

pcaobjl <- stats::prcomp(t(na.omit(assays(sumExpObj_filt)$loess)), )

percentVarl <- round(pcaobjl$sdev^2 / sum(pcaobjl$sdev^2),2)*100

pcamatl1 <- as.data.frame(pcaobjl$x)
pcamatl2 <- as.data.frame(merge(pcamatl1, colData(sumExpObj_filt), by = 0))
pcamatl2
```

### Plot sample overview using PCA

```{r eval=FALSE, include=FALSE}
plot.pca1 <- ggplot(pcamatl2, aes(PC1, PC2, color = Batch, text = Row.names)) + 
  geom_point(size=2, alpha = 0.9) + 
  scale_colour_brewer(palette = "Set2") + 
  xlab(paste("PC1: ", percentVarl[1], "% variance", sep="")) + 
  ylab(paste("PC2: ", percentVarl[2], "% variance", sep="")) + 
  theme_bw() + 
   geom_text(
    label=pcamatl2$time, 
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T
  ) + 
  ggtitle("PCA color indicating the batch and lable indicating the time-point") +
  theme(legend.title = element_blank())

#ggplotly(plot.pca12)

ggsave(plot.pca1,
  file = paste("../",
    "PCA_batch_time", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)
randomisation_code

plot.pca12 <- ggplot(pcamatl2, aes(PC1, PC2, color = Batch, text = Row.names)) + 
  geom_point(size=2, alpha = 0.9) + 
  scale_colour_brewer(palette = "Set2") + 
  xlab(paste("PC1: ", percentVarl[1], "% variance", sep="")) + 
  ylab(paste("PC2: ", percentVarl[2], "% variance", sep="")) + 
  theme_bw() + 
  geom_text(
    label=pcamatl2$randomisation_code, 
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T
  ) +
  ggtitle("PCA color indicating the batch and lable indicating the temperature group") +
  theme(legend.title = element_blank())
plot.pca12  
ggsave(plot.pca12,
  file = paste("../",
    "PCA_batch_rand_code", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)
```




```{r eval=FALSE, include=FALSE}
pca <- stats::prcomp(t(na.omit(assays(sumExpObj_filt)$loess)), )

g <- ggbiplot(pca,
  obs.scale = 1, var.scale = 1,
  groups = sumExpObj_filt$time, ellipse = T,
  circle = TRUE, var.axes = F, labels = sumExpObj_filt$Batch
) +
  scale_color_discrete(name = "") +
  theme(
    legend.direction = "horizontal",
    legend.position = "top"
  ) +
  theme_bw()

ggsave(g,
  file = paste("../Results/Figures/",
    "PCA_lables", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 15, width = 10
)


```

## Cluster samples

```{r  eval=FALSE, include=FALSE, echo = FALSE, fig.width=20, fig.height=18, warning = FALSE, message= FALSE}
distsl <- dist(t(assays(sumExpObj_filt)$loess))
plot(hclust(distsl))
```

## Explained variance by PCs

This step calculates explained variance by each principle component of the protein data.

```{r eval=FALSE, include=FALSE}
png("../Results/Figures/PC_VariancePlot.png", units = "cm", height = 24, width = 20, res = 300)
barplot(percentVarl[1:10] , ylab = "Percent varinace explained", xlab = "PC 1 to 10", main = "Percent variace explained by first 10 PCs (loess)", col = "purple")
dev.off()
```

## Correlation of principle component with Phenotype data

To identify the most relevant clinical variables associated with protein data, we calculate correlation between the principle components and the phenotype variables.

```{r eval=FALSE, include=FALSE}

pheno <- pcamatl2[ ,clin_vars]

pc_adj_rsq_l <- matrix(NA, ncol = 10, nrow = dim(pheno)[2])

for (i in 1:dim(pheno)[2]) {
  pc_adj_rsq_l[i, ] <- apply(pcamatl2[, 2:11], 2, function(x) summary(lm(x ~ pheno[, i]))$adj.r.squared)
}
colnames(pc_adj_rsq_l) <- colnames(pcamatl2[, 2:11])
rownames(pc_adj_rsq_l) <- colnames(pheno)


png("../corrPlot_updated.png", units = "cm", height = 24, width = 20, res = 300)
pheatmap::pheatmap(pc_adj_rsq_l, display_numbers = TRUE, fontsize = 12, cluster_cols = FALSE, main = "Adj R^2 of Association between PCs and clinical  and technical variables")
dev.off()
```

## PCA using VSN normalization

```{r}
res.pca_vsn <- PCA(t(assays(sumExpObj_filt)$vsn),  graph = TRUE)
```

```{r}
exp_var_vsn <- fviz_screeplot(res.pca_vsn, addlabels = TRUE, ylim = c(0, 12))
ggsave(exp_var_vsn,
  file = paste("../",
    "explained_var_vsn", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)
```


```{r, eval = FALSE,include=FALSE,  echo = FALSE, fig.width=10, fig.height=18, warning = FALSE, message= FALSE}
is.na(assays(Data_se_comp)$imputed) %>% sum()

pcaobjl_vsn <- stats::prcomp(t(na.omit(assays(sumExpObj_filt)$vsn)), )

percentVarl_vsn <- round(pcaobjl_vsn$sdev^2 / sum(pcaobjl_vsn$sdev^2),2)*100

pcamatl1_vsn <- as.data.frame(pcaobjl_vsn$x)
pcamatl2_vsn <- as.data.frame(merge(pcamatl1_vsn, colData(sumExpObj_filt), by = 0))
pcamatl2_vsn
```

### Plot sample overview using PCA

```{r eval=FALSE, include=FALSE}
plot.pca1_vsn <- ggplot(pcamatl2_vsn, aes(PC1, PC2, color = Batch, text = Row.names)) + 
  geom_point(size=2, alpha = 0.9) + 
  scale_colour_brewer(palette = "Set2") + 
  xlab(paste("PC1: ", percentVarl_vsn[1], "% variance", sep="")) + 
  ylab(paste("PC2: ", percentVarl_vsn[2], "% variance", sep="")) + 
  theme_bw() + 
   geom_text(
    label=pcamatl2$time, 
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T
  ) + 
  ggtitle("PCA color indicating the batch and lable indicating the time-point") +
  theme(legend.title = element_blank())

#ggplotly(plot.pca12)

ggsave(plot.pca1_vsn,
  file = paste("../",
    "PCA_batch_time_vsn", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)


plot.pca12_vsn <- ggplot(pcamatl2_vsn, aes(PC1, PC2, color = Batch, text = Row.names)) + 
  geom_point(size=2, alpha = 0.9) + 
  scale_colour_brewer(palette = "Set2") + 
  xlab(paste("PC1: ", percentVarl_vsn[1], "% variance", sep="")) + 
  ylab(paste("PC2: ", percentVarl_vsn[2], "% variance", sep="")) + 
  theme_bw() + 
  geom_text(
    label=pcamatl2_vsn$randomisation_code, 
    nudge_x = 0.25, nudge_y = 0.25, 
    check_overlap = T
  ) +
  ggtitle("PCA color indicating the batch and lable indicating the temperature group") +
  theme(legend.title = element_blank())
plot.pca12  
ggsave(plot.pca12_vsn,
  file = paste("../",
    "PCA_batch_rand_code_vsn", ".png",
    sep = ""
  ),
  scale = 2,
  units = "cm", height = 10, width = 15
)
```





## Correlation of principle component with Phenotype data

To identify the most relevant clinical variables associated with protein data, we calculate correlation between the principle components and the phenotype variables.

```{r eval=FALSE, include=FALSE}

pheno <- pcamatl2_vsn[ ,clin_vars]

pc_adj_rsq_l_vsn <- matrix(NA, ncol = 10, nrow = dim(pheno)[2])

for (i in 1:dim(pheno)[2]) {
  pc_adj_rsq_l_vsn[i, ] <- apply(pcamatl2_vsn[, 2:11], 2, function(x) summary(lm(x ~ pheno[, i]))$adj.r.squared)
}
colnames(pc_adj_rsq_l_vsn) <- colnames(pcamatl2_vsn[, 2:11])
rownames(pc_adj_rsq_l_vsn) <- colnames(pheno)


png("../corrPlot_updated_vsn.png", units = "cm", height = 24, width = 20, res = 300)
pheatmap::pheatmap(pc_adj_rsq_l_vsn, display_numbers = TRUE, fontsize = 12, cluster_cols = FALSE, main = "Adj R^2 of Association between PCs and clinical  and technical variables")
dev.off()
```

## Save the filtered data and PCs

```{r}
colData(sumExpObj_filt) <- merge(colData(sumExpObj_filt), pcamatl1[, 1:15], by = 0)
#saveRDS(sumExpObj_filt, "../normalized_filtered_se.rds")
```




-------------------------------

```{r setup2, echo=TRUE ,include=TRUE}

# Get Quantities of PG.ProteinGroups
nn_sp_report_l <- melt(data = nn_sp_report, id.vars = "PG.ProteinGroups", measure.vars = na.omit(str_match(colnames(nn_sp_report), '.*?PG.Quantity'))[,1], variable.name = 'sample')

# table of missing values count
nn_missing <- nn_sp_report_l %>% distinct(`PG.ProteinGroups`,sample,value) %>% group_by(sample) %>% mutate(sum_na = sum(is.nan(value))) %>% ungroup() %>% 
  group_by(`PG.ProteinGroups`) %>% mutate(sum_na_prot = sum(is.nan(value))) %>% mutate(sum_pep = length(`PG.ProteinGroups`)) %>% dplyr::select(`PG.ProteinGroups`,sample,sum_na, sum_na_prot, sum_pep)

# CK imputation method
max.impute.value <- nn_sp_report_l %>%
    group_by(`PG.ProteinGroups`) %>%
    summarise(min_int = min(value, na.rm = T)) %>%
    mutate(max_imp = 0.5*min_int)
nn_sp_report_l <- nn_sp_report_l %>%
    left_join(max.impute.value) %>%
    dplyr::rowwise() %>%
    mutate(Imputed = if_else(is.nan(value),TRUE, FALSE)) %>%
    mutate(value = if_else(is.nan(value), runif(1,min = 0.005*max_imp, max = max_imp), value))

# log2 of value (intensity)
nn_sp_report_l <- nn_sp_report_l %>% mutate(log2_val = log2(value))   

# Calculate protein groups quantitative intensity (quant_int) and also log10, select unique protein - sample relation  
nn_sp_report_l <- nn_sp_report_l %>% group_by(`PG.ProteinGroups`, sample) %>% mutate(quant_int = mean(value), quant_int_log10 = log10(quant_int)) %>% ungroup() %>% distinct(`PG.ProteinGroups`,sample,quant_int,quant_int_log10,Imputed)

# Normalize by subtraction of mean 
nn_sp_report_l <- nn_sp_report_l %>% group_by(sample) %>% mutate(scale_log10 = scale(quant_int_log10))

# Add timepoint column by splitting the sample identifier
nn_sp_report_l <- dplyr::rowwise(nn_sp_report_l) %>% mutate("time" = str_match(sample, "-[0-9]{2}-"), 
                                                            "sample" =  str_split(str_split(sample, ".htrms.")[[1]][1], " ")[[1]][2]) 
# Get total count of proteins (1 protein count per protein group)
total_pro <- nn_sp_report_l %>% dplyr::select(`PG.ProteinGroups`) %>% ungroup() %>% distinct(`PG.ProteinGroups`) %>% add_count(., name = "proteins")

# Make new table for peptide and protein counts
total_all <- cbind(total_pro[1,"proteins"])
colnames(total_all) <- c('proteins')
ta <- melt(data = total_all, measure.vars = colnames(total_all), variable.name = "id")



```


```{r setup3, echo=TRUE}
# Use RSQLite to count proteins
db <- dbConnect(RSQLite::SQLite(),"")
dbWriteTable(db,"nn_sp", nn_sp_report_l)
dbListTables(db)
nn_sp_report_counted <- dbGetQuery(db, 'SELECT `PG.ProteinGroups`,nn_sp.sample,quant_int,quant_int_log10,scale_log10,time,protein_count,Imputed FROM nn_sp LEFT JOIN(SELECT sample,count(Imputed) AS protein_count FROM nn_sp WHERE Imputed == 0 GROUP BY sample) AS pt ON pt.sample = nn_sp.sample')

dbDisconnect(db)

```

Run normalizeVSN from limma:
```{r normalizeVSN, echo=TRUE}
db <- dbConnect(RSQLite::SQLite(),"")
df_w <- dcast(nn_sp_report_l, `PG.ProteinGroups` ~ sample,value.var = "scale_log10")
rownames(df_w) <- df_w[,1]

df_imp <- dcast(nn_sp_report_l, `PG.ProteinGroups` ~ sample,value.var = "Imputed")
rownames(df_imp) <- df_imp[,1]
df_w[which(df_imp==TRUE)]

df_w <- as.matrix(df_w[,-1])
df_imp <- as.matrix(df_imp[,-1])
df_w_norm <- limma::normalizeVSN(df_w[,-1])
df_l_norm <- melt(data = df_w_norm, id.vars = rownames(df_w_norm), measure.vars = colnames(df_w_norm), variable.name = "sample" )
df_l_norm <- df_l_norm %>% rename("norm_log10" = "value")
nn_sp_report_counted <- left_join(nn_sp_report_counted, df_l_norm, by = c("PG.ProteinGroups"="Var1","sample"="Var2"))

```
Choose Random samples for plotting:
```{r sample_samples, echo=TRUE}
result_table <- nn_sp_report_counted
# distributions of sample of samples
samples <- result_table %>% distinct(sample) %>% as.list()
s = sample(samples$sample, 16)

```


```{r subsets_samples, echo=TRUE} 
# function for breaking down samples in subsets 
subset_samples <- function(sample_list, step){
  count <- 0
  subset_list <- list()
  for (i in seq(1,length(sample_list), by=step)){
    count <- count + 1
    subset_list[[count]] <- sample_list[i:(i+step-1)]
    
    }
return(subset_list)
}

# distributions of sample of samples
samples <- result_table %>% distinct(sample) %>% as.list()
s = sample(samples$sample, 16)
subset_list <- subset_samples(samples$sample, 16)
```


Write dataset output:
```{r write_output, echo=TRUE}
write_tsv(nn_sp_report_counted,"/home/rstudio/data/01_quality_check_table_directDIA.tsv" )

```


Plot total counts and counts by sample:
```{r total_counts, echo=TRUE, fig.height=6, fig.width=8}
# Plot total counts
title <- "Total counts"
total_count  <- ggplot(ta %>% 
                         rename("count" = "value","proteins" = "id"), 
                         aes(proteins,count)) +
  geom_bar(position = "dodge",stat = "identity") +
  theme(axis.text.x= element_text(angle = 90, vjust = 1, hjust=1 ,size = 8)) + 
  geom_text(aes(label = count), size = 4, color = "white", vjust= 1.5) +
  ggtitle(paste0(title)) 
(total_count)

# protein count by sample
title <- "Protein intensity by count"
plt_prot_dot_z <- ggplot(result_table %>% 
                        group_by(sample) %>% 
                        mutate(mean_quant_int = mean(quant_int)), 
                        aes(protein_count, mean_quant_int, colour = time)) +
                        geom_point() +
  theme(axis.text.x= element_text(angle = 0, vjust = 1, hjust=0.5 ,size = 10)) + 
  ggtitle(paste0(title)) 
(plt_prot_dot_z)

```

Boxplots to visualize sample variation before and after normalization:  
```{r plot2, echo=TRUE,  fig.height=6, fig.width=8}

# Boxplot before normalization 
title <- "Before normalization: Box plot of log10"
plt_box <- ggplot(result_table %>% filter(sample %in% s), aes(sample, quant_int_log10, fill = time)) +
  geom_boxplot() +
  theme(axis.text.x= element_text(angle = 90, vjust = 1, hjust=1 ,size = 7)) +
  ggtitle(paste0(title)) 
plot(plt_box)

# Boxplot after normalization 
title <- "After normalization: Box plot of log10"
plt_box <- ggplot(result_table %>% filter(sample %in% s), aes(sample, scale_log10, fill = time)) +
  geom_boxplot() +
  theme(axis.text.x= element_text(angle = 90, vjust = 1, hjust=1 ,size = 7)) + 
  ggtitle(paste0(title)) 
plot(plt_box)

 

```

Histograms to see how imputation affects distribution: 
```{r plot4, echo=TRUE, fig.height=6, fig.width=8}

  title <- "Without imputation: log10 protein intensities"
  plt_hist <- ggplot(result_table %>% filter(sample %in% s, Imputed == FALSE), aes(x = scale_log10)) +
    geom_histogram(bins = 60) +
    facet_wrap(~ sample) +
    ggtitle(paste0(title))
  plot(plt_hist)

  title <- "With imputation: log10 protein intensities"
  plt_hist <- ggplot(result_table %>% filter(sample %in% s), aes(x = scale_log10, fill = as.factor(Imputed))) +
    geom_histogram(bins = 60) +
     
    facet_wrap(~ sample) +
    ggtitle(paste0(title))
  plot(plt_hist)


```
QQ-plots to see how imputed values affect normal distribution:

```{r plot5, echo=TRUE, fig.height=6, fig.width=8}


  # qq-plots of sample of samples 
  title <- "Before imputation: QQ-plot of samples"
  plt_qq <- ggplot(result_table %>% filter(sample %in% s, Imputed == FALSE),aes( sample = scale_log10,  ), na.rm = T) +
    geom_qq() +
    facet_wrap(~ sample) +
    ggtitle(paste0(title))
  plot(plt_qq)

  title <- "After imputation : QQ-plot of samples"
  plt_qq <- ggplot(result_table %>% filter(sample %in% s),aes( sample = scale_log10  ), na.rm = T) +
    geom_qq() +
    facet_wrap(~ sample) +
    ggtitle(paste0(title))
  plot(plt_qq)


``` 

