---
title: "Winter batch - Ovarian Cancer, case control analyses"
author: "Ashfaq Ali"
date: "2022-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup1, echo=FALSE ,include=FALSE}
# import libraries and data
#renv::init()
library(tidyverse)
library(reshape2)
library(magrittr)
library(readxl)
library(fuzzyjoin)
library(Hmisc)
library(limma)
library(DBI)
library(VIM)
library(factoextra)
library(FactoMineR)
library(mice)
library(SummarizedExperiment)
#renv::install("stemangiola/tidySummarizedExperiment")
library(xlsx)
#library(tidySummarizedExperiment)

library(RColorBrewer)
library(GGally)
library(ggiraphExtra)
library(knitr)
library(kableExtra)
library(plotly)
library(RankAggreg)
library(DT)
library(ggVennDiagram)
getwd()

source("./code/functions.R")
```

# Re-analyses with case control categories

```{r}
sumExpObj_overall <- readRDS( "./data/rawdata/sumExpObj_overall_wnt_filt.rds")

```

## Filter and prune outcome variable

Samples corresponding the "other" category were removed along with the samples with NA values in the outcome variable.

Based on the meeting with Karin in Spring, the outcome variable corresponding to tumor type were pruned as following.

For 0,1,2 coding, 1 and 2 were converted into one level.

For tumor type variable, "B" and "H" categories were not combined to form "Control" group while "BL" and "M" were combined to make the "case" category.


```{r}
sumExpObj_overall <- readRDS( "./data/rawdata/sumExpObj_overall_wnt_filt.rds")

sumExpObj_winter_overall_filt <- sumExpObj_overall[, !is.na(sumExpObj_overall$Type_1_2) | !is.na(sumExpObj_overall$tumor_type) ]
sumExpObj_winter_overall_filt$tumor_type[grep("Other", sumExpObj_winter_overall_filt$tumor_type)] <- "Other"
sumExpObj_winter_overall_filt <- sumExpObj_winter_overall_filt[ , which(sumExpObj_winter_overall_filt$tumor_type!="Other") ]

sumExpObj_winter_overall_filt$tumor_type_new <- case_when(
sumExpObj_winter_overall_filt$tumor_type =="B" |  sumExpObj_winter_overall_filt$tumor_type == "H" ~ "control",
  sumExpObj_winter_overall_filt$tumor_type =="BL" |  sumExpObj_winter_overall_filt$tumor_type == "M" ~ "case"
                                     
                                        )


sumExpObj_winter_overall_filt$tumor_type_new <- as.factor(sumExpObj_winter_overall_filt$tumor_type_new) |> relevel(ref = "control")

sumExpObj_winter_overall_filt$Type_1_2_new <- case_when(
sumExpObj_winter_overall_filt$Type_1_2 ==0 ~ "One",
  sumExpObj_winter_overall_filt$Type_1_2 ==1 |  sumExpObj_winter_overall_filt$Type_1_2 == 2 ~ "Two"
                                     
                                        )

sumExpObj_winter_overall_filt$Type_1_2_new  <- as.factor(sumExpObj_winter_overall_filt$Type_1_2_new ) |> relevel(ref = "One")


saveRDS(sumExpObj_winter_overall_filt, "./data/rawdata/data_sumExpfil_wntr_new_vars.rds")

```



## Statistical ananlyses: Overall data set for Loess Normalized data

### type

```{r}

#design_t12 <- model.matrix(~ 0 + Type_1_2, data = colData(sumExpObj_summer_overall_filt))

design_tt <- model.matrix(~ 0 + tumor_type_new, data = colData(sumExpObj_winter_overall_filt))

name <- levels(sumExpObj_winter_overall_filt$tumor_type_new)
colnames(design_tt) <- name

fit_tt <- lmFit((assays(sumExpObj_winter_overall_filt)$batch_rm_l_raw_loess), design = design_tt)
contrast.matrix <- makeContrasts(case-control, levels=name)
fit_tt <- contrasts.fit(fit_tt, contrast.matrix)
fit_tt <- eBayes(fit = fit_tt)


Results_overall_wntr <- list()

Results_overall_wntr$case_vs_control <- topTable(fit_tt, p.value = 1, number = Inf, coef = "case - control", confint = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column(., var = "Protein.Group") %>%
 # arrange(padj) %>%
  left_join(rowData(sumExpObj_winter_overall_filt) |> as_tibble(),
    {
      .
    },
    by = "Protein.Group"
  ) 



```

### Tumor type variable

```{r}
design_tt2 <- model.matrix(~ 0 + Type_1_2_new, data = colData(sumExpObj_winter_overall_filt))

name <- levels(sumExpObj_winter_overall_filt$Type_1_2_new)
#name <- c("zero", "one", "two")
colnames(design_tt2) <- name

fit_12 <- lmFit((assays(sumExpObj_winter_overall_filt)$batch_rm_l_raw_loess), design = design_tt2)
contrast.matrix <- makeContrasts(Two-One, levels=name)
fit_12 <- contrasts.fit(fit_12, contrast.matrix)
fit_12 <- eBayes(fit = fit_12)


#Results_overall <- list()

Results_overall_wntr$type_two_vs_on <- topTable(fit_12, p.value = 1, number = Inf, coef = "Two - One", confint = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column(., var = "Protein.Group") %>%
 # arrange(padj) %>%
  left_join(rowData(sumExpObj_summer_overall_filt) |> as_tibble(),
    {
      .
    },
    
    by ="Protein.Group"
  )
```


```{r}
writexl::write_xlsx(Results_overall_wntr, "./results_wnt/DE_results_winter.xlsx")

```

